### 2023-11-22 chapters/discussion.tex
Work on discussion

\moretodo[inline]{Add to the discussion somewhere:
I would try to embed a perspective for contemplation. Instead of us always talking about what the technology needs to improve on. Perhaps, it would also be prudent to ask an open-ended philosophical question: “While our requirements to the technology are in its place, one could ask, whether human beings have the ability to quantify uncertainty. In 2023, the prevalence of human errors in driving are much higher than self-driving vehicles.” <- Quick example. I'm sure you can find something better.}
%

The rapid progress of machine learning that started about a decade ago with the 2012 ImageNet competition and the work of \textcite{krizhevsky_imagenet_2012} all but slowed down during the course of this project. 
Following the papers by \textcite{choi_waic_2019,nalisnick_detecting_2019,hendrycks_deep_2019}, the field of out-of-distribution detection saw a surge of interest that has since grown yearly, while self"=supervised learning for speech has continued to advance with works such as CPC \parencite{oord_representation_2018}, wav2vec 2.0 \parencite{baevski_wav2vec_2020} and data2vec \parencite{baevski_data2vec_2022}.

@@

Two key observations might lead to the pessimistic view that VAEs are fundamentally unsuited for representation learning. 
First, VAEs are trained to maximize the ELBO which is a marginal likelihood objective conditionally independent of the latent variables $\zb$. Second, as we saw in \cref{eq:variational-autoencoder-elbo-factorized-expected}, the ELBO contains can be written to reveal that its maximization leads to minimization of the mutual information between the observed and latent variables. In the following we will discuss each of these observations.

@@

On the other hand, when we use maximum likelihood to train a latent variable model, we maximize the log-marginal likelihood $\log p(\xb)$. This can be equivalently phrased as maximizing the KL-divergence of the true data distribution $p_{\text{data}}(\xb)$ to the learned marginal, $D_{\mathrm{KL}}(p_{\text{data}}(\xb)\parallel p_{\thetab}(\xb))$\footnote{%
Maximizing the likelihood $\sum_{n=1}^N \log p_\theta(\xb)$ w.r.t. parameters $\thetab$ is asymptotically equivalent to minimizing $D_{\mathrm{KL}}(p_{\text{data}}(\xb)\parallel p_{\thetab}(\xb))$: 
%
\begin{align}
    \argmin_\thetab D_{\mathrm{KL}}(p_{\text{data}}(\xb)\parallel p_{\thetab}(\xb)) 
    &= \argmin_\thetab \mathbb{E}_{p_{\text{data}}(\xb)} \left[ \log \frac{p_{\text{data}}(\xb)}{p_{\theta}(\xb)} \right] \nonumber \\
    % &= \argmin_\thetab \mathbb{E}_{p_{\text{data}}(\xb)} \left[ \log p_{\text{data}}(\xb) - \log p_{\theta}(\xb) \right] \nonumber \\
    &= \argmax_\thetab \mathbb{E}_{p_{\text{data}}(\xb)} \left[ \log p_{\theta}(\xb) \right] \nonumber \\
    &= \argmax_\thetab \lim_{n\rightarrow\infty} \underset{\text{log-likelihood}}{\underbrace{\sum_{n=1}^N \log p_{\theta}(\xb_n)}} \enspace . \nonumber
\end{align}
%
The last equality follows from the law of large numbers.
% \begin{align}
%     \mathcal{L}(\thetab; p_{\mathrm{data}}(\xb))
%     &= \lim_{N\rightarrow\infty} \frac{1}{N} \sum\limits_{n=1}^{N} \log p_{\thetab}(\xb) \\
%     &= \mathbb{E}_{p_{\mathrm{data}}(\xb)} \left[ \log p_{\thetab}(\xb) \right] \\
%     &= \mathbb{E}_{p_{\mathrm{data}}(\xb)} \left[ \log p_{\thetab}(\xb) - \log p_{\mathrm{data}}(\xb) + \log p_{\mathrm{data}}(\xb) \right] \\
%     &= \mathbb{E}_{p_{\mathrm{data}}(\xb)} \left[ \log p_{\mathrm{data}}(\xb) \right] + \mathbb{E}_{p_{\mathrm{data}}(\xb)} \left[ \log\frac{ p_{\thetab}(\xb)}{p_{\mathrm{data}}(\xb) }\right] \\
%     &= - \mathbb{H}[p_{\mathrm{data}}(\xb)] - D_\mathrm{KL}\left( p_{\mathrm{data}}(\xb)\mid\mid p(\xb) \right)
% \end{align}
}. 
The first key observation then is that the marginal $p_{\thetab}(\xb)$, which we maximize, and the posterior $p_{\thetab}(\zb|\xb)$, which we want to produce useful representations, are separate properties of a latent variable model. Any combination uniquely defines a valid latent variable model. Hence, maximizing the marginal is a completely useless criteria for representation learning, regardless of the measure of usefulness \parencite{huszar_is_2017, alemi_fixing_2018}.

\paragraph{Mutual information minimization}
As we saw in \cref{subsec: mutual-information-interpretation-of-elbo}, optimizing the ELBO involves minimizing the mutual information between the latent and observed variables. We reprint equation \cref{eq:variational-autoencoder-elbo-factorized-expected} below for reference.
%
\begin{equation} %\label{eq:variational-autoencoder-elbo-factorized-expected (discussion copy)}
    \mathbb{E}_{q_{\phib}(\xb,\zb)} \left[ \mathcal{L}_{}(\xb;\thetab,\phib) \right] = 
    \underset{\text{average reconstruction}}{\underbrace{
        \mathbb{E}_{q_{\phib}(\xb,\zb)}\left[ \log p_{\thetab}(\xb | \zb) \right]
    }}
    - 
    \underset{\text{marginal KL to prior}}{\underbrace{
    \vphantom{\mathbb{E}_{q_{\phib}(\xb,\zb)}\left[ \log p_{\thetab}(\xb | \zb) \right]}{
        D_{\mathrm{KL}}\left[ q_{\phib}(\zb) \parallel p(\zb) \right]
    }
    }} 
    -
    \underset{\text{mutual information}}{\underbrace{
    I_{q_{\phib}(\xb,\zb)}\left[\xb ; \zb \right]
    }} \. . \nonumber
\end{equation}
%
Note that we let $q_{\phib}(\xb,\zb) = q_{\phib}(\zb|\xb)\hat{p}(\xb)$ for ease of notation. 
Optimizing the ELBO can be understood as simultaneously maximizing the average reconstruction likelihood, minimizing the marginal KL-divergence to the prior, and minimizing the mutual information between the observed and latent variables.

\parencite{tomczak_trouble_2022}

\moretodo[inline]{Finish the discussion of the MI interpretation}

@@

In practice, the maximum marginal likelihood estimation takes place not over all possible models, but over a constrained parametric class defined by the inference network $q_\thetab(\zb|\xb) \approx p(\zb|\xb)$ and the generative model $p_\thetab(\xb|\zb)p(\zb) = p_\theta(\xb,\zb)$. 
This introduces a coupling between the marginal $p_{\thetab}(\xb)$ and the posterior $p_{\thetab}(\zb|\xb)$ which makes the maximum likelihood solution unlikely to correspond to completely useless representations \parencite{huszar_is_2017}. 
Similarly, maximizing the ELBO in \cref{eq:variational-autoencoder-elbo-factorized-expected (discussion copy)} by exactly minimizing the mutual information term corresponds to a complete posterior collapse that requires a trade-off with the other terms. This solution is almost surely a local minimum and is often avoided in practice.
The challenge is then how to impose appropriate constraints to make VAEs learn representations that are competitive with self"=supervised methods.

@@

Including masked objectives into the principled probabilistic framework of VAEs might require creative thinking, but it seems likely that relaxing the requirement of exact input reconstruction and enforcing the need of encoding (and decoding) contextual information could lead the way for future types of VAEs to learn representations that are competitive with self"=supervised methods.

@@

% If the stroke recognition ensemble were to be deployed in a randomized controlled trial, s
Calibrated probabilities in the alerts presented to users might be a way to alleviate alarm fatigue. Calibrated probabilities would allow users to discern between certain and uncertain predictions and also enable the system to present users only with predictions that have at least a certain probability of being correct. Similar approaches have been suggested by clinicians and interviews indicate that predictive uncertainty is perceived by experts as a sort of explanation that complements the prediction \cite{tonekaboni_what_2019}. In \cref{fig_discussion:retrospective-paper-f1-performance-vs-predicted-probability} we show the F1-score of the ensemble model and the call-takers computed on subsets of the data created by binning the calibrated probabilities. We note that, as might be expected, both call-taker and ensemble model performance increase with increased model certainty. This indicates that selecting, based on certainty, which predictions to present to users might indeed help build trust in the system and ensure a practical impact.

>>>>

\moretodo[inline]{Add to the discussion somewhere:
I would try to embed a perspective for contemplation. Instead of us always talking about what the technology needs to improve on. Perhaps, it would also be prudent to ask an open-ended philosophical question: “While our requirements to the technology are in its place, one could ask, whether human beings have the ability to quantify uncertainty. In 2023, the prevalence of human errors in driving are much higher than self-driving vehicles.” <- Quick example. I'm sure you can find something better.}

The rapid progress of machine learning that started about a decade ago with the 2012 ImageNet competition and the work of \textcite{krizhevsky_imagenet_2012} all but slowed down during the course of this project. 
Following the papers by \textcite{choi_waic_2019,nalisnick_detecting_2019,hendrycks_deep_2019}, the field of out-of-distribution detection saw a surge of interest that has since grown yearly, while representation learning for speech has continued to advance with notable works such as CPC \parencite{oord_representation_2018}, wav2vec 2.0 \parencite{baevski_wav2vec_2020} and data2vec \parencite{baevski_data2vec_2022}.

@@

Two key observations might lead to the pessimistic view that VAEs are fundamentally unsuited for representation learning. 
First, VAEs are trained to maximize the ELBO which includes a marginal likelihood objective that is independent of the latent variables $\zb$. Second, as we saw in \cref{eq:variational-autoencoder-elbo-factorized-expected}, the ELBO can be written to reveal that its maximization leads to minimization of the mutual information between the observed and latent variables. In the following we will discuss each of these observations.

@@

On the other hand, when we use maximum likelihood to train a latent variable model, we maximize the log-marginal likelihood $\log p(\xb)$. This can be equivalently phrased as maximizing the KL-divergence of the true data distribution $p_{\text{data}}(\xb)$ to the learned marginal, $D_{\mathrm{KL}}(p_{\text{data}}(\xb)\parallel p_{\thetab}(\xb))$\footnote{%
Maximizing the likelihood $\sum_{n=1}^N \log p_\theta(\xb)$ w.r.t. parameters $\thetab$ is asymptotically equivalent to minimizing $D_{\mathrm{KL}}(p_{\text{data}}(\xb)\parallel p_{\thetab}(\xb))$: 
%
\begin{align}
    \argmin_\thetab D_{\mathrm{KL}}(p_{\text{data}}(\xb)\parallel p_{\thetab}(\xb)) 
    &= \argmin_\thetab \mathbb{E}_{p_{\text{data}}(\xb)} \left[ \log \frac{p_{\text{data}}(\xb)}{p_{\theta}(\xb)} \right] \nonumber \\
    % &= \argmin_\thetab \mathbb{E}_{p_{\text{data}}(\xb)} \left[ \log p_{\text{data}}(\xb) - \log p_{\theta}(\xb) \right] \nonumber \\
    &= \argmax_\thetab \mathbb{E}_{p_{\text{data}}(\xb)} \left[ \log p_{\theta}(\xb) \right] \nonumber \\
    &= \argmax_\thetab \lim_{n\rightarrow\infty} \underset{\text{log-likelihood}}{\underbrace{\sum_{n=1}^N \log p_{\theta}(\xb_n)}} \enspace . \nonumber
\end{align}
%
The last equality follows from the law of large numbers.
% \begin{align}
%     \mathcal{L}(\thetab; p_{\mathrm{data}}(\xb))
%     &= \lim_{N\rightarrow\infty} \frac{1}{N} \sum\limits_{n=1}^{N} \log p_{\thetab}(\xb) \\
%     &= \mathbb{E}_{p_{\mathrm{data}}(\xb)} \left[ \log p_{\thetab}(\xb) \right] \\
%     &= \mathbb{E}_{p_{\mathrm{data}}(\xb)} \left[ \log p_{\thetab}(\xb) - \log p_{\mathrm{data}}(\xb) + \log p_{\mathrm{data}}(\xb) \right] \\
%     &= \mathbb{E}_{p_{\mathrm{data}}(\xb)} \left[ \log p_{\mathrm{data}}(\xb) \right] + \mathbb{E}_{p_{\mathrm{data}}(\xb)} \left[ \log\frac{ p_{\thetab}(\xb)}{p_{\mathrm{data}}(\xb) }\right] \\
%     &= - \mathbb{H}[p_{\mathrm{data}}(\xb)] - D_\mathrm{KL}\left( p_{\mathrm{data}}(\xb)\mid\mid p(\xb) \right)
% \end{align}
}. 
The first key observation then is that the marginal $p_{\thetab}(\xb)$, which we maximize, and the posterior $p_{\thetab}(\zb|\xb)$, which we want to produce useful representations, are separate properties of a latent variable model. Any combination uniquely defines a valid latent variable model. Hence, maximizing the marginal is a useless criterion for representation learning, regardless of the measure of usefulness \parencite{huszar_is_2017, alemi_fixing_2018}.

However, VAEs are not trained to maximize the marginal likelihood directly, but a lower bound of it, the ELBO,
%
\begin{equation} \label{eq:variational-autoencoder-elbo-factorized (discussion copy)}
    \mathcal{L}_{}(\xb;\thetab,\phib)
    = 
    \underset{\text{reconstruction loss}}{\underbrace{
        \mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right]
    }}
     - 
    \underset{\text{KL-divergence to prior}}{
        \underbrace{\vphantom{\mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right]}{D_\text{KL}\left( q_\phib(\zb|\xb) \parallel p(\zb) \right)}
    }} \enspace .
\end{equation}
%
Since the ELBO is not independent of the latent variables, its maximization will typically depend on the latent variables and the posterior $p(\zb|\xb)$. Hence, unlike the marginal likelihood, the ELBO is not guaranteed to be a useless criterion for representation learning. Nonetheless, being a lower bound of the marginal likelihood that we seek to maximize, seems to indicate that this issue might not be irrelevant to consider.

\paragraph{Mutual information minimization}
As we saw in \cref{subsec: mutual-information-interpretation-of-elbo}, optimizing the ELBO involves minimizing the mutual information between the latent and observed variables. We reprint equation \cref{eq:variational-autoencoder-elbo-factorized-expected} below for reference.
%
\begin{equation} \label{eq:variational-autoencoder-elbo-factorized-expected (discussion copy)}
    \mathbb{E}_{q_{\phib}(\xb,\zb)} \left[ \mathcal{L}_{}(\xb;\thetab,\phib) \right] = 
    \underset{\text{average reconstruction}}{\underbrace{
        \mathbb{E}_{q_{\phib}(\xb,\zb)}\left[ \log p_{\thetab}(\xb | \zb) \right]
    }}
    - 
    \underset{\text{marginal KL to prior}}{\underbrace{
    \vphantom{\mathbb{E}_{q_{\phib}(\xb,\zb)}\left[ \log p_{\thetab}(\xb | \zb) \right]}{
        D_{\mathrm{KL}}\left[ q_{\phib}(\zb) \parallel p(\zb) \right]
    }
    }} 
    -
    \underset{\text{mutual information}}{\underbrace{
    I_{q_{\phib}(\xb,\zb)}\left[\xb ; \zb \right]
    }} \. . %\nonumber
\end{equation}
%
Note that we let $q_{\phib}(\xb,\zb) = q_{\phib}(\zb|\xb)\hat{p}(\xb)$ for ease of notation. 
% Optimizing the ELBO can be understood as simultaneously maximizing the average reconstruction likelihood, minimizing the marginal KL-divergence to the prior, and minimizing the mutual information between the observed and latent variables. 
We can see that optimizing the ELBO means learning a good one-to-one mapping of $z$ to $x$ (average reconstruction), fitting the aggregated posterior to the prior (marginal KL-divergence), while also not learning any stochastic dependency between $x$ and $z$ (mutual information). 
These objectives are conflicting and especially the minimization of mutual information seems counterproductive to our goals \parencite{tomczak_trouble_2022}.

@@

In practice, the maximum marginal likelihood estimation takes place not over all possible models, but over a constrained parametric class defined by the inference network $q_\thetab(\zb|\xb) \approx p(\zb|\xb)$ and the generative model $p_\thetab(\xb|\zb)p(\zb) = p_\theta(\xb,\zb)$. 
This introduces a coupling between the marginal $p_{\thetab}(\xb)$ and the posterior $p_{\thetab}(\zb|\xb)$ which makes the maximum likelihood solution unlikely to correspond to completely useless representations \parencite{huszar_is_2017}. 
Similarly, maximizing the ELBO in \cref{eq:variational-autoencoder-elbo-factorized-expected (discussion copy)} by exactly minimizing the mutual information term corresponds to a complete posterior collapse that requires a trade-off with the other terms. For a given model, this solution is almost surely a local minimum, but one that is often avoided in practice. 
The challenge is then how to impose appropriate constraints to make VAEs learn representations that are competitive with self"=supervised methods.

@@

Including masked objectives into the principled probabilistic framework of VAEs might require rethinking the principles themselves, but it seems likely that relaxing the requirement of exact input reconstruction and enforcing the need for encoding (and decoding) contextual information could pave the way for future types of VAEs to learn representations that are competitive with self"=supervised methods.

@@

% If the stroke recognition ensemble were to be deployed in a randomized controlled trial, s
Calibrated probabilities in the alerts presented to users might be a way to alleviate alarm fatigue. Calibrated probabilities would allow users to discern between certain and uncertain predictions and also enable the system to present users only with predictions that have a minimum probability of being correct. Similar approaches have been suggested by clinicians and interviews indicate that predictive uncertainty is perceived by experts as a sort of explanation that complements the prediction \cite{tonekaboni_what_2019}. In \cref{fig_discussion:retrospective-paper-f1-performance-vs-predicted-probability} we show the F1-score of the ensemble model and the call-takers computed on subsets of the data created by binning the calibrated probabilities. We note that, as might be expected, both call-taker and ensemble model performance increase with increased model certainty. This indicates that selecting, based on certainty, which predictions to present to users might indeed help build trust in the system and ensure a practical impact.
